---
title: "FIRE"
author: "Tali Mass"
date: "2025-05-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Install and load packages

```{r load_packages}
library(tidyverse)
#install.packages("lubridate")
library(broom)
library(lubridate)
#install.packages("cowplot")
library(cowplot)
library(tidyverse)
#install.packages("ggtext")
library(ggtext)
#install.packages("rstatix")
library(rstatix)
library(car)
#install.packages("FSA")
library(FSA)
#install.packages("rcompanion")
library(rcompanion)
library(ggpubr)
library(dplyr)
library(vegan)
library(ggplot2)
library(RColorBrewer)
library(car)
library(tidyr)
library(emmeans)
library(reshape2)
library(scales)
library(ARTool)
library(multcomp)
library(rcompanion)
library(psych)
```

#Load Data
```{r}

fire.data <- read.csv(
  "/Users/talimass/Documents/Documents - MacBook Pro/GitHub/Costar-Rica-TPC/FIRE_all_sites_anotated.xls.csv",
  stringsAsFactors = FALSE
)
```

# View Data plots
fvfm = Quantum yield of photochemistry in PSII
Pmax = max photosynthesis   
sigma = Functional absorption cross-section of PSII
p = conectivity 
Fm = Max Fluorescent
Qa = reoxidation kinetics in dark

```{r}

  

# ==========================================
# Universal plotting function (Facet Species)
# ==========================================

plot_psii_species <- function(data, variable, y_label, file_name) {
  
  # Ensure numeric
  data[[variable]] <- as.numeric(data[[variable]])
  data$Temp <- as.numeric(data$Temp)
  
  # Remove NA safely
  data_clean <- data %>%
    filter(!is.na(.data[[variable]]),
           !is.na(Temp),
           !is.na(Site),
           !is.na(Species))
  
  # Remove outliers (IQR method)
  Q1 <- quantile(data_clean[[variable]], 0.25)
  Q3 <- quantile(data_clean[[variable]], 0.75)
  IQR_value <- Q3 - Q1
  
  lower <- Q1 - 1.5 * IQR_value
  upper <- Q3 + 1.5 * IQR_value
  
  data_filtered <- data_clean %>%
    filter(.data[[variable]] >= lower &
           .data[[variable]] <= upper)
  
  cat("\n---------------------------------\n")
  cat("Variable:", variable, "\n")
  cat("Rows before filtering:", nrow(data_clean), "\n")
  cat("Rows after filtering :", nrow(data_filtered), "\n")
  
  # Plot
  p <- ggplot(data_filtered,
              aes(x = factor(Temp),
                  y = .data[[variable]],
                  fill = Site)) +
    
    geom_boxplot(alpha = 0.6,
                 outlier.shape = NA,
                 position = position_dodge(width = 0.8)) +
    
    geom_jitter(aes(color = Site),
                position = position_jitterdodge(
                  jitter.width = 0.15,
                  dodge.width = 0.8
                ),
                size = 1.5,
                alpha = 0.6) +
    
    facet_wrap(~ Species) +
    
    theme_classic() +
    
    labs(title = paste(variable, "across Temperature by Species"),
         x = "Temperature",
         y = y_label) +
    
    theme(legend.position = "right")
  
  print(p)
  
  # Save
  output_folder <- "/Users/talimass/Documents/Documents - MacBook Pro/GitHub/Costar-Rica-TPC/Output"
  dir.create(output_folder, showWarnings = FALSE)
  
  ggsave(
    filename = file.path(output_folder, file_name),
    plot = p,
    width = 10,
    height = 6,
    dpi = 300
  )
}

# ==========================================
# Generate All Plots
# ==========================================

plot_psii_species(fire.data, "fv_fm",
                  "Fv/Fm",
                  "fvfm_species.png")

plot_psii_species(fire.data, "Sigma",
                  expression(sigma[PSII]),
                  "Sigma_species.png")

plot_psii_species(fire.data, "Pmax.e.s",
                  expression(P[max]~"(electrons s"^{-1}~"PSII"^{-1}*")"),
                  "Pmax_species.png")

plot_psii_species(fire.data, "p",
                  "p (–)",
                  "p_species.png")

plot_psii_species(fire.data, "Fm",
                  "Fm",
                  "Fm_species.png")

plot_psii_species(fire.data, "TauAv1",
                  expression(tau[Av1]),
                  "TauAv1_species.png")


```


#FIRe STATS
#Kruskal-Wallis Test- data is not normally distributed

```{r}
str(fire.data)
levels(fire.data$Temp)
levels(fire.data$Species)
```
# in case you need to select a specific species use the row below. 
```{r}
fire.past=fire.data[fire.data$Species=="Pavona",]
str(fire.past)
```
# We subset the columns we should test for. at any time, you can add or retract from this list. NOTE, you should look at the specific manner by which the name of the variable is in the object itself.
```{r}
test.vars=c("Temp", "Site", "fv_fm", "Sigma", "p", "Pmax.e.s")

fire.data.to.test=fire.data[,names(fire.data) %in% test.vars]
str(fire.data.to.test)

```
# test for normality and homogeneity of variance
```{r}
Spiro_test=data.frame(do.call(rbind, lapply(fire.data.to.test[,3:6], function(x) shapiro.test(x)[c("statistic", "p.value")])))
Spiro_test

Leven_test=lapply(fire.data.to.test[,3:6], leveneTest, group = fire.data.to.test$Temp)
Leven_test
```
# if any or all the test are significant, you will need to use non-parametric tests. if both are not significant, you can use ANOVA.


### in case the data is normally distributed and the variance is homogeneous


```{r}
#One-way ANOVA test
aov_fire.data <- apply(fire.data.to.test[,3:6], 2, function(x) aov(x ~ Temp, data = fire.data.to.test))
aov_res_df <- as.data.frame(do.call(rbind, lapply(aov_fire.data, broom::tidy)))
aov_res_df

```
## post-hoc Tukey test, only in case ANOVA was significant, for pair-wise comparisons
```{r}
tukey_res <- sapply(aov_fire.data , function(x) TukeyHSD(x, "Temp", ordered = TRUE))
tukey_res_df <- as.data.frame(do.call(rbind, Map(cbind, Name = names(tukey_res), tukey_res)))
tukey_res_df

```
### in case the data is not normally distributed and/or variance is not equal

##Kruskal-Walis test
```{r}
KW_fire.data <- apply(fire.data.to.test[,3:6], 2, function(x) kruskal.test(x ~ Temp, data = fire.data.to.test))
KW_res_df <- as.data.frame(do.call(rbind, lapply(KW_fire.data, broom::tidy)))
KW_res_df
```
##post-hoc Dunn test, in case any of KW test are significant, for pairwise comparisons.
```{r}
Dunn_fire.data <- apply(fire.data.to.test[,3:6], 2, function(x) dunnTest(x ~ Temp, data = fire.data.to.test, table=TRUE))
Dunn_res_df_Fv.Fm <- data.frame(Dunn_fire.data$Fv.Fm$res)
Dunn_res_df_Fv.Fm$Factor="Fv.Fm"
Dunn_res_df_Sigma <- data.frame(Dunn_fire.data$Sigma$res)
Dunn_res_df_Sigma$Factor="Sigma"
Dunn_res_df_p <- data.frame(Dunn_fire.data$p$res)
Dunn_res_df_p$Factor="p"
Dunn_res_df_Pmax.e.s <- data.frame(Dunn_fire.data$Pmax.e.s$res)
Dunn_res_df_Pmax.e.s$Factor="Pmax.e.s"

Dunn_all=rbind(Dunn_res_df_Fv.Fm,Dunn_res_df_Sigma,Dunn_res_df_p,Dunn_res_df_Pmax.e.s)
```

