---
title: "FIRE"
author: "Tali Mass"
date: "2025-05-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Install and load packages

```{r load_packages}
library(tidyverse)
#install.packages("lubridate")
library(broom)
library(lubridate)
#install.packages("cowplot")
library(cowplot)
library(tidyverse)
#install.packages("ggtext")
library(ggtext)
#install.packages("rstatix")
library(rstatix)
library(car)
#install.packages("FSA")
library(FSA)
#install.packages("rcompanion")
library(rcompanion)
library(ggpubr)
library(dplyr)
library(vegan)
library(ggplot2)
library(RColorBrewer)
library(car)
library(tidyr)
library(emmeans)
library(reshape2)
library(scales)
library(ARTool)
library(multcomp)
library(rcompanion)
library(psych)
#install.packages("tinytex")
#tinytex::install_tinytex()

```

#Load Data
```{r}

fire.data <- read.csv("/Users/talimass/Documents/Documents - MacBook Pro/GitHub/Costar-Rica-TPC/FIRE_Samara_anotated.xls.csv", stringsAsFactors = TRUE)




fire.data$Site_depth <- paste(fire.data$Site, fire.data$depth, sep = " - ")
fire_data <- fire.data %>% na.omit()
```

# View Data plots
fvfm = Quantum yield of photochemistry in PSII
Pmax = max photosynthesis   
sigma = Functional absorption cross-section of PSII
p = conectivity 
Fm = Max Fluorescent
Qa = reoxidation kinetics in dark


```{r}
str(fire_data$fv_fm)
class(fire_data$fv_fm)
# Convert character or factor to numeric
fire_data$fv_fm <- as.numeric(as.character(fire_data$fv_fm))

# Now compute quartiles
quartiles <- quantile(fire_data$fv_fm, probs = c(0.25, 0.75), na.rm = TRUE)
IQR_cor <- IQR(fire_data$fv_fm)

lower <- quartiles[1] - 1.5*IQR_cor
upper <- quartiles[2] + 1.5*IQR_cor

fvfm_no_outlier <- subset(fire_data, fire_data$fv_fm >lower & fire_data$fv_fm< upper)
#plot of Each species is in a separate panel using

fvfm <- ggplot(fvfm_no_outlier,
               aes(x = factor(Temp), y = fv_fm, fill = Species)) +
  geom_boxplot() +
  stat_boxplot(geom = "errorbar", width = 0.75) +
  theme_classic() +
  labs(title = "Quantum Yield of PSII by Species",
       x = "Temperature",
       y = "Fv/Fm") +
  theme(legend.position = "none") +
  facet_wrap(~ Species)

print(fvfm)


# Define the file name and path for saving
output_folder <- "/Users/talimass/Documents/Documents - MacBook Pro/GitHub/Costar-Rica-TPC/Output"
file_name <- paste0(output_folder, "/fvfm.png")

# Save the plot
ggsave(filename = file_name, plot = fvfm, width = 8, height = 6, dpi = 300)

# Print the saved file location
print(paste("Saved the combined plot at", file_name))

# Plot with boxplots and all data points (scatter) on top
fvfm_plot <- ggplot(data = fvfm_no_outlier, 
                    aes(x = factor(Temp), y = fv_fm, fill = Species)) +
  
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +   # Boxplots
  geom_jitter(color = "black", width = 0.15, size = 1, alpha = 0.6) +  # Scatter points
  stat_boxplot(geom = "errorbar", width = 0.5) +    # Error bars
  
  theme_classic() +
  
  labs(title = "Quantum Yield of PSII by Species",
       x = "Temperature",
       y = "Fv/Fm") +
  
  theme(legend.position = "none") +
  
  facet_wrap(~ Species, ncol = 2)

# Display plot
print(fvfm_plot)

# Define the file name and path for saving
file_name <- paste0(output_folder, "/fvfm_plot.png")

# Save the plot
ggsave(filename = file_name, plot = fvfm_plot, width = 8, height = 6, dpi = 300)

# Print the saved file location
print(paste("Saved the combined plot at", file_name))



```

#Sigma

```{r}
#quartiles <- quantile(fire_data$Sigma, probs = c(.25,.75))

str(fire_data$Sigma)
class(fire_data$Sigma)
# Convert character or factor to numeric
fire_data$Sigma <- as.numeric(as.character(fire_data$Sigma))


# Ensure Sigma is numeric
fire_data$Sigma <- as.numeric(as.character(fire_data$Sigma))

# Calculate quartiles (25th and 75th percentiles)
quartiles <- quantile(fire_data$Sigma, probs = c(0.25, 0.75), na.rm = TRUE)

# Calculate IQR (Interquartile Range)
IQR_cor <- IQR(fire_data$Sigma, na.rm = TRUE)

# Calculate lower and upper bounds for outliers
lower <- quartiles[1] - 1.5 * IQR_cor
upper <- quartiles[2] + 1.5 * IQR_cor

# Subset the data to remove outliers based on Sigma
sigma_no_outlier <- subset(fire_data, Sigma > lower & Sigma < upper)

# Create the plot

sigma_plot <- ggplot(sigma_no_outlier,
               aes(x = factor(Temp), y = Sigma, fill = Species)) +
  geom_boxplot() +
  stat_boxplot(geom = "errorbar", width = 0.75) +
  theme_classic() +
  labs(title = "Functional Absorption Cross-section of PSII",
       x = "Temperature",
       y = "σPSII’(A2)") +
  theme(legend.position = "none") +
  facet_wrap(~ Species)

print(sigma_plot)





# Save the plot
output_folder <- "/Users/talimass/Documents/Documents - MacBook Pro/GitHub/Costar-Rica-TPC/Output"
file_name <- paste0(output_folder, "/sigma_plot.png")

# Use the correct plot object (sigma_plot)
ggsave(filename = file_name, plot = sigma_plot, width = 8, height = 6, dpi = 300)

# Print the saved file location
print(paste("Saved the plot at", file_name))


#plot with all data points (scatter points) on top of the boxplots
# Boxplot with all data points (scatter) on top
sigma_plot <- ggplot(data = sigma_no_outlier, 
                     aes(x = factor(Temp), y = Sigma, fill = Species)) +
  
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(color = "black", width = 0.15, size = 1, alpha = 0.6) +
  stat_boxplot(geom = "errorbar", width = 0.5) +
  
  theme_classic() +
  
  labs(title = "Functional Absorption Cross-section of PSII",
       x = "Temperature",
       y = "σPSII’(A2)") +  
  
  theme(legend.position = "none") +
  
  facet_wrap(~ Species, ncol = 2)

print(sigma_plot)


# Define the file name and path for saving
file_name <- paste0(output_folder, "/sigma.png")



# Save the plot
ggsave(filename = file_name, plot = sigma_plot, width = 8, height = 6, dpi = 300)

# Print the saved file location
print(paste("Saved the combined plot at", file_name))

```
```{r}
quartiles <- quantile(fire_data$Pmax.e.s, probs = c(.25,.75))
IQR_cor <- IQR(fire_data$Pmax.e.s)

lower <- quartiles[1] - 1.5*IQR_cor
upper <- quartiles[2] + 1.5*IQR_cor

Pmax.e.s_no_outlier <- subset(fire_data, fire_data$Pmax.e.s >lower & fire_data$Pmax.e.s< upper)


# Create the plot


pmax_plot <- ggplot(Pmax.e.s_no_outlier,
               aes(x = factor(Temp), y = Pmax.e.s, fill = Species)) +
  geom_boxplot() +
  stat_boxplot(geom = "errorbar", width = 0.75) +
  theme_classic() +
  labs(title = "Quantum Yield of PSII by Species",
       x = "Temperature",
       y = "Pmax (electron s-1 PSII-1)") +
  theme(legend.position = "none") +
  facet_wrap(~ Species)

print(pmax_plot)



# Save the plot
output_folder <- "/Users/talimass/Documents/Documents - MacBook Pro/GitHub/Costar-Rica-TPC/Output"
file_name <- paste0(output_folder, "/pmax_plot.png")

# Use the correct plot object (pmax_plot)
ggsave(filename = file_name, plot = pmax_plot, width = 8, height = 6, dpi = 300)

# Print the saved file location
print(paste("Saved the plot at", file_name))

#plot with all data points (scatter points) on top of the boxplots

pmax_plot <- ggplot(Pmax.e.s_no_outlier,
                    aes(x = factor(Temp), y = Pmax.e.s, fill = Species)) +
  
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(color = "black", width = 0.15, size = 1, alpha = 0.6) +
  stat_boxplot(geom = "errorbar", width = 0.5) +
  
  theme_classic() +
  
  labs(title = "Maximum Electron Transport Rate (Pmax)",
       x = "Temperature",
       y = expression(P[max]~"(electrons s"^{-1}~"PSII"^{-1}*")")) +
  
  theme(legend.position = "none") +
  
  facet_wrap(~ Species)

print(pmax_plot)
 # Define the file name and path for saving
file_name <- paste0(output_folder, "/pmax.png")

# Save the plot
ggsave(filename = file_name, 
       plot = pmax_plot, 
       width = 8, height = 6, dpi = 300)


# Print the saved file location
print(paste("Saved the combined plot at", file_name))

```
```{r}
str(fire_data$p)
class(fire_data$p)
# Convert character or factor to numeric
fire_data$p <- as.numeric(as.character(fire_data$p))

# Now compute quartiles
quartiles <- quantile(fire_data$p, probs = c(0.25, 0.75), na.rm = TRUE)
IQR_cor <- IQR(fire_data$p)

lower <- quartiles[1] - 1.5*IQR_cor
upper <- quartiles[2] + 1.5*IQR_cor


p_no_outlier <- subset(fire_data, fire_data$p >lower & fire_data$p< upper)



# Create the plot

p_plot <- ggplot(p_no_outlier,
               aes(x = factor(Temp), y = p, fill = Species)) +
  geom_boxplot() +
  stat_boxplot(geom = "errorbar", width = 0.75) +
  theme_classic() +
  labs(title = "Conectivity",
       x = "Temperature",
       y = "p") +
  theme(legend.position = "none") +
  facet_wrap(~ Species)

print(p_plot)



# Save the plot
output_folder <- "/Users/talimass/Documents/Documents - MacBook Pro/GitHub/Costar-Rica-TPC/Output"
file_name <- paste0(output_folder, "/p_plot.png")

# Use the correct plot object (pmax_plot)
ggsave(filename = file_name, plot = p_plot, width = 8, height = 6, dpi = 300)

# Print the saved file location
print(paste("Saved the plot at", file_name))

#plot with all data points (scatter points) on top of the boxplots

p_plot <- ggplot(p_no_outlier,
                 aes(x = factor(Temp), y = p, fill = Species)) +
  
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(color = "black", width = 0.15, size = 1, alpha = 0.6) +
  stat_boxplot(geom = "errorbar", width = 0.5) +
  
  theme_classic() +
  
  labs(title = "Connectivity",
       x = "Temperature",
       y = "p (–)") +
  
  theme(legend.position = "none") +
  
  facet_wrap(~ Species)

print(p_plot)

# Define the file name and path for saving
file_name <- paste0(output_folder, "/p.png")

# Save the plot
ggsave(filename = file_name, plot = p_plot, width = 8, height = 6, dpi = 300)

# Print the saved file location
print(paste("Saved the combined plot at", file_name))
```

```{r}
str(fire_data$Fm)
class(fire_data$Fm)
#Convert character or factor to numeric
fire_data$Fm <- as.numeric(as.character(fire_data$Fm))
sum(is.na(fire_data$Fm))

# Now compute quartiles
quartiles <- quantile(fire_data$Fm, probs = c(0.25, 0.75), na.rm = TRUE)
IQR_cor <- IQR(fire_data$Fm)

lower <- quartiles[1] - 1.5*IQR_cor
upper <- quartiles[2] + 1.5*IQR_cor

Fm_no_outlier <- subset(fire_data, fire_data$Fm >lower & fire_data$Fm< upper)


# Create the plot


Fm_plot <- ggplot(Fm_no_outlier,
               aes(x = factor(Temp), y = Fm, fill = Species)) +
  geom_boxplot() +
  stat_boxplot(geom = "errorbar", width = 0.75) +
  theme_classic() +
  labs(title = "Maximum Fluorescence (Fm)",
       x = "Temp",
       y = "Fm") +
  theme(legend.position = "none") +
  facet_wrap(~ Species)

print(Fm_plot)



# Save the plot
output_folder <- "/Users/talimass/Documents/Documents - MacBook Pro/GitHub/Costar-Rica-TPC/Output"
file_name <- paste0(output_folder, "/Fm_plot.png")

# Use the correct plot object (Fm_plot)
ggsave(filename = file_name, plot = Fm_plot, width = 8, height = 6, dpi = 300)

# Print the saved file location
print(paste("Saved the plot at", file_name))

#plot with all data points (scatter points) on top of the boxplots

Fm_plot <- ggplot(Fm_no_outlier,
                  aes(x = factor(Temp), y = Fm, fill = Species)) +
  
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(color = "black", width = 0.15, size = 1, alpha = 0.6) +
  stat_boxplot(geom = "errorbar", width = 0.5) +
  
  theme_classic() +
  
  labs(title = "Maximum Fluorescence (Fm)",
       x = "Temperature",
       y = "Fm") +
  
  theme(legend.position = "none") +
  
  facet_wrap(~ Species)

print(Fm_plot)

# Define the file name and path for saving
file_name <- paste0(output_folder, "/Fm.png")

# Save the plot
ggsave(filename = file_name, plot = Fm_plot, width = 8, height = 6, dpi = 300)

# Print the saved file location
print(paste("Saved the combined plot at", file_name))

```


```{r}
str(fire_data$TauAv1)
class(fire_data$TauAv1)
#Convert character or factor to numeric
fire_data$TauAv1 <- as.numeric(as.character(fire_data$TauAv1))
sum(is.na(fire_data$TauAv1))

# Now compute quartiles
quartiles <- quantile(fire_data$TauAv1, probs = c(0.25, 0.75), na.rm = TRUE)
IQR_cor <- IQR(fire_data$TauAv1)

lower <- quartiles[1] - 1.5*IQR_cor
upper <- quartiles[2] + 1.5*IQR_cor

TauAv1_no_outlier <- subset(fire_data, fire_data$TauAv1 >lower & fire_data$TauAv1< upper)


# Create the plot



TauAv1_plot <- ggplot(TauAv1_no_outlier,
               aes(x = factor(Temp), y = TauAv1, fill = Species)) +
  geom_boxplot() +
  stat_boxplot(geom = "errorbar", width = 0.75) +
  theme_classic() +
  labs(title = "Qa reoxidation kinetics in dark",
       x = "Temperature",
       y = "TauAv1") +
  theme(legend.position = "none") +
  facet_wrap(~ Species)

print(TauAv1_plot)


# Save the plot
output_folder <- "/Users/talimass/Documents/Documents - MacBook Pro/GitHub/Costar-Rica-TPC/Output"
file_name <- paste0(output_folder, "/TauAv1_plot.png")

# Use the correct plot object (TauAv1_plot)
ggsave(filename = file_name, plot = TauAv1_plot, width = 8, height = 6, dpi = 300)

# Print the saved file location
print(paste("Saved the plot at", file_name))

#plot with all data points (scatter points) on top of the boxplots

TauAv1_plot <- ggplot(TauAv1_no_outlier,
                      aes(x = factor(Temp), y = TauAv1, fill = Species)) +
  
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(color = "black", width = 0.15, size = 1, alpha = 0.6) +
  stat_boxplot(geom = "errorbar", width = 0.5) +
  
  theme_classic() +
  
  labs(title = "QA Reoxidation Kinetics in Dark",
       x = "Temperature",
       y = expression(tau[Av1])) +
  
  theme(legend.position = "none") +
  
  facet_wrap(~ Species)

print(TauAv1_plot)

# Define the file name and path for saving
file_name <- paste0(output_folder, "/TauAv1.png")

# Save the plot
ggsave(filename = file_name, plot = TauAv1_plot, width = 8, height = 6, dpi = 300)

# Print the saved file location
print(paste("Saved the combined plot at", file_name))

```
#FIRe STATS
#Kruskal-Wallis Test- data is not normally distributed

```{r}
str(fire.data)
fire.data$Species <- as.factor(fire.data$Species)
fire.data$Temp <- as.factor(fire.data$Temp)

```

# We subset the columns we should test for. at any time, you can add or retract from this list. NOTE, you should look at the specific manner by which the name of the variable is in the object itself.
```{r}
test.vars <- c("fv_fm", "Sigma", "p", "Pmax.e.s", "Fm", "TauAv1")


fire.data.to.test=fire.data[,names(fire.data) %in% test.vars]
str(fire.data.to.test)

```
```{r setup, include=FALSE}
options(encoding = "UTF-8")
knitr::opts_chunk$set(echo = TRUE)


```


# test for normality and homogeneity of variance
```{r}



output_folder <- "/Users/talimass/Documents/Documents - MacBook Pro/GitHub/Costar-Rica-TPC/Output"

Shapiro_species <- lapply(test.vars, function(var) {
  
  fire.data %>%
    group_by(Species, Temp) %>%
    summarise(
      statistic = shapiro.test(.data[[var]])$statistic,
      p_value   = shapiro.test(.data[[var]])$p.value,
      .groups = "drop"
    ) %>%
    mutate(Variable = var)
})

Shapiro_species_df <- bind_rows(Shapiro_species)

write.table(
  Shapiro_species_df,
  file = file.path(output_folder, "Shapiro_results.txt"),
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)







```
# if any or all the test are significant, you will need to use non-parametric tests. if both are not significant, you can use ANOVA.


### in case the data is normally distributed and the variance is homogeneous



```{r Levene_test, echo=FALSE, results='hide', message=FALSE, warning=FALSE}



output_folder <- "/Users/talimass/Documents/Documents - MacBook Pro/GitHub/Costar-Rica-TPC/Output"
dir.create(output_folder, recursive = TRUE, showWarnings = FALSE)

fire.data$Species <- as.factor(fire.data$Species)
fire.data$Temp    <- as.factor(fire.data$Temp)

test.vars <- c("fv_fm", "Sigma", "p", "Pmax.e.s", "Fm", "TauAv1")

Levene_results_list <- list()

for (var in test.vars) {

  if (!var %in% colnames(fire.data)) next

  for (sp in unique(fire.data$Species)) {

    sub_data <- subset(fire.data, Species == sp)

    # Need at least 2 temperature levels
    if (length(unique(sub_data$Temp)) < 2) next

    test_result <- tryCatch(
      leveneTest(as.formula(paste(var, "~ Temp")), data = sub_data),
      error = function(e) NULL
    )

    if (!is.null(test_result)) {
      temp_df <- as.data.frame(test_result)
      temp_df$Species  <- sp
      temp_df$Variable <- var
      Levene_results_list[[length(Levene_results_list) + 1]] <- temp_df
    }
  }
}

Levene_species_df <- do.call(rbind, Levene_results_list)

sink(file.path(output_folder, "Levene_results.txt"))

cat("=====================================\n")
cat("LEVENE HOMOGENEITY TEST RESULTS\n")
cat("=====================================\n\n")

if (is.null(Levene_species_df) || nrow(Levene_species_df) == 0) {
  cat("No valid Levene test results generated.\n")
} else {
  print(Levene_species_df)
}

sink()

```

# if any or all the test are significant, you will need to use non-parametric tests. if both are not significant, you can use ANOVA.


### in case the data is normally distributed and the variance is homogeneous




```{r}
#One-way ANOVA test
aov_fire.data <- apply(fire.data.to.test[,3:6], 2, function(x) aov(x ~ Temp, data = fire.data.to.test))
aov_res_df <- as.data.frame(do.call(rbind, lapply(aov_fire.data, broom::tidy)))
aov_res_df

```
## post-hoc Tukey test, only in case ANOVA was significant, for pair-wise comparisons
```{r}
tukey_res <- sapply(aov_fire.data , function(x) TukeyHSD(x, "Temp", ordered = TRUE))
tukey_res_df <- as.data.frame(do.call(rbind, Map(cbind, Name = names(tukey_res), tukey_res)))
tukey_res_df

```
### in case the data is not normally distributed and/or variance is not equal

##Kruskal-Walis test per species
```{r}




output_folder <- "/Users/talimass/Documents/Documents - MacBook Pro/GitHub/Costar-Rica-TPC/Output"
dir.create(output_folder, recursive = TRUE, showWarnings = FALSE)

fire.data$Species <- as.factor(fire.data$Species)
fire.data$Temp    <- as.factor(fire.data$Temp)

test.vars <- c("fv_fm", "Sigma", "p", "Pmax.e.s", "Fm", "TauAv1")

results_list <- list()

for (var in test.vars) {
  
  for (sp in unique(fire.data$Species)) {
    
    sub_data <- subset(fire.data, Species == sp)
    
    # Remove NA for that variable
    sub_data <- sub_data[!is.na(sub_data[[var]]), ]
    
    # Must have at least 2 temperature groups
    if (length(unique(sub_data$Temp)) < 2) next
    
    test <- kruskal.test(as.formula(paste(var, "~ Temp")), data = sub_data)
    
    results_list[[length(results_list)+1]] <- data.frame(
      Variable = var,
      Species  = sp,
      Chi_sq   = test$statistic,
      df       = test$parameter,
      p_value  = test$p.value
    )
  }
}

Kruskal_species_df <- do.call(rbind, results_list)

write.table(
  Kruskal_species_df,
  file = file.path(output_folder, "Kruskal_results.txt"),
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)

```



##post-hoc Dunn test, in case any of KW test are significant, for pairwise comparisons.
```{r}



# Output folder
output_folder <- "/Users/talimass/Documents/Documents - MacBook Pro/GitHub/Costar-Rica-TPC/Output"
dir.create(output_folder, recursive = TRUE, showWarnings = FALSE)

fire.data$Species <- as.factor(fire.data$Species)
fire.data$Temp    <- as.factor(fire.data$Temp)

test.vars <- c("fv_fm", "Sigma", "p", "Pmax.e.s", "Fm", "TauAv1")

Dunn_results_list <- list()

for (var in test.vars) {
  
  for (sp in unique(fire.data$Species)) {
    
    sub_data <- subset(fire.data, Species == sp)
    
    # Remove NA
    sub_data <- sub_data[!is.na(sub_data[[var]]), ]
    
    # Need at least 2 temperature groups
    if (length(unique(sub_data$Temp)) < 2) next
    
    # Run Dunn test
    dunn_res <- tryCatch(
      dunnTest(as.formula(paste(var, "~ Temp")),
               data = sub_data,
               method = "bonferroni"),  # change to "holm" or "BH" if preferred
      error = function(e) NULL
    )
    
    if (!is.null(dunn_res)) {
      
      temp_df <- dunn_res$res
      temp_df$Variable <- var
      temp_df$Species  <- sp
      
      Dunn_results_list[[length(Dunn_results_list) + 1]] <- temp_df
    }
  }
}

Dunn_species_df <- do.call(rbind, Dunn_results_list)

# Save results
write.table(
  Dunn_species_df,
  file = file.path(output_folder, "Dunn_results.txt"),
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)

```




